\scshape Evolutionary dynamics, equilibrium selection, and what population and quantitative genetics have taught evolutionary game theory

### 2022-11-08 manuscript.tex
now onto examples

\usepackage{setspace}
\usepackage{titling}
\usepackage{xparse}
\usepackage{lineno}

@@

\newcommand{\der}{\mathop{}\!\mathrm{d}}
\newcommand{\Oh}[1]{O\left(#1\right)}
\newcommand{\ExpectationOperator}{\mathrm{E}}
\NewDocumentCommand \E {d[] d()} {%
  \IfNoValueTF {#1} {%
    \IfNoValueTF {#2} {%
      \mathop{\kern0pt\ExpectationOperator}\nolimits%
    }{%
      \mathop{\kern0pt\ExpectationOperator}\nolimits\left(#2\right)%
    }}{%
    \IfNoValueTF {#2} {%
      \mathop{\kern0pt\ExpectationOperator#1}\nolimits%
    }{%
      \mathop{\kern0pt\ExpectationOperator#1}\nolimits\left(#2\right)%
    }%
  }
}
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\Cov}{Cov}
\newcommand{\Fst}{F_{\mathrm{ST}}}
\newcommand{\mean}[1]{\overline{#1}}
\newcommand{\w}{w}
\newcommand{\ess}[1]{#1^*}
\newcommand{\fixp}[1]{\hat{#1}}
\renewcommand{\vec}[1]{\symbf{#1}}
\newcommand{\rec}{\rho}
\newcommand{\mut}{\mu}
\newcommend{\eig}{\lambda}

@@

The role of evolutionary game theory and the ESS method specifically in evolutionary biology has been at the crux of some of the most important conceptual debates in the field including the role of natural selection and adaptation vis-a-vis other forces \cite[e.g.,][]{MaynardSmith:1978,Gould:Lewontin:1979,Lewontin:1979,Orzack:Sober:1994,Gardner:2017,Kern:Hahn:2018,Jensen:Payseur:2019} and the importance of kin and group selection relative to individual selection
\cite[e.g.,][]{Maynard-Smith:1964,Hamilton:1963,Price:1972:cov,Wilson:Wilson:2007,Leigh:2010,Akcay:VanCleve:2012,West:Griffin:2007,Gardner:Grafen:2009,Nowak:Tarnita:2010,Abbot:Abe:2011,Allen:Nowak:2013,Birch:2014,Birch:2017,Nowak:McAvoy:2017}. These debates might be captured in part by the following two questions: \begin{ilnum} \item \label{q:I} How do ESS models that focus on individual fitness capture the effects of kin selection or group selection? \item \label{q:II} How does an ESS account for mutation, genetic drift, and other evolutionary forces? \end{ilnum} In this work, I will describe how conceptual advances in evolutionary theory since Maynard Smith and Price \cite{Maynard-Smith:Price:1973} have shed light on both of these questions and have led to a broader understanding of the ESS method and a more integrative framework for understanding how multiple evolutionary forces in complex populations can lead to diverse phenotypes.

Question \ref{q:I} derives in part from how Maynard Smith introduced his work on ESSs; he argued that an ESS could provide an explanation for the evolution of behaviors where ``selection acts entirely at the individual level, but in which the success of any particular strategy depends on what strategies are adopted by other members of the population'' \cite[][p. 210]{Maynard-Smith:1974} and is not ``due to group or species selection'' \cite[p. 15]{Maynard-Smith:Price:1973} or selection on ``close relatives'' \cite[p. 210]{Maynard-Smith:1974} as might be the case for kin selection \cite{Hamilton:1964}. In setting up this dichotomy, Maynard Smith implies that the ESS method may not apply when kin or group selection are involved and that kin or group selection models might lead to different results when applied to the same biological scenario. As I will describe below, evolutionary theorists have since realized (including Maynard Smith himself \cite[p. 33]{MaynardSmith:1978}) that the ESS method is really orthogonal to issues of individual vs. group vs. kin selection and instead captures in what sense evolution via natural selection optimizes fitness given a specific measure of fitness. Issues of individual vs. group vs. kin selection are about the ``units'' or ``levels'' at which selection acts and how fitness should be measured to account for selection at those levels. Lewontin also hints at the levels of selection question in his 1961 game theory paper when he discusses the relative merits of individual-level measures of fitness like intrinsic or Malthusian growth rate $r$ versus population-level measures like mean fitness $\mean{w}$ \cite[p. 400-401]{Lewontin:1961} as analogs of utility. The levels of selection question became a major topic of study in evolutionary theory \cite[e.g.,][]{Lewontin:1970,Dawkins:1982,Wilson:Sober:1989,Maynard-Smith:Szathmary:1995,Wilson:1997,Michod:1999,Michod:2006,Szathmary:2015} and philosophy of biology \cite{Hull:1980,Brandon:1982,Damuth:Heisler:1988,Lloyd:1992,Lloyd:1994,Sober:Wilson:1994,Okasha:2006,Okasha:2016} and continues to generate substantial research \cite[e.g.,][]{Black:Bourrat:2020,Cooney:Mori:2022,Veit:2022}. For the purpose of explaining how an ESS can capture selection at multiple levels and among relatives or kin, I will argue below that the right measure of utility is the ``lineage fitness'' \cite{Lehmann:Alger:2015,Akcay:VanCleve:2016,Lehmann:Mullon:2016} of a mutant allele at a single genetic locus, which can be shown to be functionally equivalent to a measure of inclusive fitness \cite{Lehmann:Mullon:2016,Lehmann:Rousset:2020}.

@@

The definition of an ESS is relatively simple and only involves a measure of fitness, a set of phenotypes, and a stability condition. Let $\w(x,y)$ be the fitness that an individual with phenotype or strategy $x$ obtains when interacting with an individual with phenotype or strategy $y$. The phenotypes can be drawn from a set of discrete or continous set of values $X$ that represent a single trait like body size or propensity to cooperate. Fitness $\w$ for the moment is simply the survival rate (i.e., viability selection) but as we'll see we can define much more general fitness measures. The stability condition \cite{Maynard-Smith:Price:1973,Maynard-Smith:1974} says that a phenotype $\ess{x}$ is an ESS, or is evolutionarily stable (ES), when no alternative strategy $y$ can improve upon the fitness that $\ess{x}$ receives when interacting with itself:
\begin{subequations}
  \label{eq:ess:def}
\begin{equation}
  \label{eq:ess:def:a}
  \w(\ess{x}, \ess{x}) \ge \w(y, \ess{x})
\end{equation}
for all phenotypes $y$. In the case that the strategy $y$ gets the same fitness interacting with $\ess{x}$ as $\ess{x}$ does with itself (equation \eqref{eq:ess:def:a} holds as an equality) then $\ess{x}$ must do better interacting with $y$ than $y$ does with itself:
\begin{equation}
  \label{eq:ess:def:b}
  \w(\ess{x}, y) > \w(y, y)
\end{equation}
\end{subequations}
The two conditions in \eqref{eq:ess:def} are equivalent to requiring that phenotype $\ess{x}$ receives higher average fitness than any alternative phenotype $y$ when $y$ is rare in the population \cite{Maynard-Smith:1974,Bishop:Cannings:1976}. If condition \eqref{eq:ess:def} holds as a strict inequality, then $\ess{x}$ is a ``strict'' ESS.

@@

is the frequency of phenotype 1 among mutant individuals once the invasion process has stabilized. Note that since the mutant allele is rare, mutant individuals can have only one mutant chromosome; the other chromosome has no mutant alleles at locus $A$ and remains at the internal equilibrium frequency $\fixp{p}_{kl}$. A necessary condition for the equilibrium with phenotype $\fixp{x}$ to be externally stable is that $\eig \le 1$ or from equation \eqref{eq:lambda:twolocus}

@@

  w(x_{\mut}, \fixp{x}) \le w(\fixp{x}, \fixp{x})

@@

for all possible mutant $A_{\mut}$ that generate mutant phenotype $x_{\mut}$. This condition is quite close to the ESS condition in \eqref{eq:ess:def} and can be used to prove that the following result for randomly mating diploid populations with two-locus genetic systems \cite{Eshel:Feldman:1984,Eshel:1996,Eshel:Feldman:1998}:

@@

\section{Invasion and lineage fitness and ESSs}

@@

While Result \ref{res:twolocus} emphasizes the role of ESSs and phenotypic evolution in the long-term process, the external stability approach used to prove \ref{res:twolocus} reveals a more general viewpoint where ESSs can depend in important ways on genetic and demographic parameters of a population. The external stability analysis assumes that the population is at some genetic equilibrium and is subject to potential invasion by a new mutant allele at low frequency and looks for a genetic equilibrium and its resulting phenotype that resists invasion by any mutant allele within a set of phenotypic effects. Before the mathematical machinery of external stability was developed, Hamilton appreciated the conceptual importance of uninvadable phenotypes in his study of the sex ratio, and he called sex ratios that resisted invasion ``unbeatable'' phenotypes \cite{Hamilton:1967}. As we saw above, determining whether a phenotype can be unbeatable and its genetic equilibria externally stable involves calculating the magnitude of the eigenvalue $\eig$ associated with the linear dynamical system that approximates the dynamics of the mutant allele frequency. Assuming that $\eig >0$, the phenotype $\ess{x}$ is unbeatable whenever $\eig(\ess{x},y) < 1$ for all $y \ne \ess{x}$ (within some set of effects).

%% a few things:
%% \item have to say \eig(x,x) = 1
%% \item have to work with strict ESS
%% \item maybe have to say this is for continuum of pure strategies?
%% \item maybe I can dodge that?
%% \item

external stability analysis assumes that new mutant allele arises in a population at low enough frequency that so that its dynamics are well approximated by a linear approximation of the genotype frequency dynamical equations. These linear dynamical systems either grow or contract geometrically depending on the magnitude of the eigenvalue $\eig$ of the (Jacobian) matrix where growth occurs for $\eig > 1$ (assuming $\eig>0$) and contraction for $\eig < 1$. Thus, rare mutant alleles will either successfully invade and reach some nonzero frequency, after which point their dynamics are no longer linear, or they will go extinct. The magnitude of the eigenvalue $\eig$ measures the geometric growth rate (e.g., equation \eqref{eq:lambda:twolocus}) of the mutant genotype relative to the resident genotypes and so is really a measure of the fitness of the invading mutant genotype and is often called ``invasion fitness'' \cite{Metz:Nisbet:1992,Heino:Metz:1998} (relatedly, ``invasion exponent'' \cite{Rand:Wilson:1994}). Regardless of the complexity of the genetics, demography, or ecology of a population, we can in principle perform an external stability analysis\footnote{This is also referred to as an ``evolutionary invasion analysis''.} that uses invasion fitness to predict the fate of rare mutants.

ESS phenotypes of the long-term process are produced by externally stable genetic equilibria, and the conditions for the stability of those equilibria depend on specific assumptions about the genetic system of the evolving population (e.g., random mating and autosomal loci).

A deeper aspect of how genotypic and phenotypic processes are enmeshed in Result \ref{res:twolocus} comes from looking at how external stability generates the ESS condition, or how phenotypic evolutionary stability emerges out of genotypic evolutionary stability.

Given a measure of invasion fitness, we can then look for a phenotype $x$ that cannot be invaded by any other phenotype;
Hamilton called such phenotypes ``unbeatable'' \cite{Hamilton:1967}.
An unbeatable phenotype $\ess{x}$ is one where all possible mutant phenotypes $y$ have an invasion fitness $\eig(y,\ess{x}) < 1$.
In fact, an unbeatable phenotype $\ess{x}$ is a strict ESSs \cite{Eshel:Feldman:1998}. To see this, recall that from equation \eqref{eq:lambda:twolocus} that the invasion fitness is just the ratio of the mutant fitness against the resident to the resident fitness against itself, or $\eig(y,\ess{x}) = w(y,\ess{x})/w(\ess{x}, \ess{x})$. Since $\eig < 1$ for all mutant $y$ phenotypes, this means $w(y,\ess{x}) < w(\ess{x},\ess{x})$, which is the definition of a strict ESS from equation \eqref{eq:ess:def:a}. Thus, finding an ESS $\ess{x}$ means calculating the right invasion fitness for a population and showing that $\eig(\ess{x},y) < 1$ for all possible $y$.

As we saw in section \ref{sec:long-term}, the invasion fitness $\eig$ has an associated eigenvector $v$ that measures the frequency of mutant genotypes during the invasion process.

If those assumptions change, then the ES phenotypes may depend on parameters of the genetic system. For example, the ES sex ratio \cite{Fisher:1930} in a randomly mating population is 1:1 when the ratio is determined by an diploid autosomal gene \cite{Hamilton:1967,Eshel:Feldman:1982}, but it can take other values when the gene is sex-linked or the genetic system is more complex like in insect that are haplodiploid \cite{Hamilton:1967,Eshel:Feldman:1982a}. A more recent example involves the evolution of bet-hedging where organisms alternative between different phenotypes with a specific probability or rate when the environmental cycles between different conditions \cite{Seger:Brockmann:1987,Slatkin:1974}. These phenotypic shifts can be due to genetic mutations \cite{Leigh:1970,Ishii:Matsuda:1989} or switching between different epigenetic states  \cite{Kussell:Leibler:2005,Salathe:VanCleve:2009} where control of the mutation or switching rate is determined by a ``modifier'' locus separate for the genetic or epigenetic locus controlling the phenotype. Due to the the fitness effects of a changing environment, this scenario does not fit neatly into the conditions required for Result \ref{res:twolocus}. In fact, an analysis of the external stability of the switching/mutation rate shows that the switching rate that is ES depends on the recombination rate between phenotypic locus and the modifier locus \cite{Liberman:VanCleve:2011}. For example, Figure \ref{fig:pip:switching}, shows pairwise invasibility plots for switching rate with different recombination rates and given weakly asymmetric selection in two different environments that alternate every $50$ generations. White regions are where mutant alleles with a switching rate on the veritcal axis can invade a population with a resident rate for the rest of the population given by the horizontal axis; black regions are when such a mutant cannot invade. The plots show that a rate of $~1/50$ is ES when the recombination rate is zero (similar to older results in \cite{Leigh:1970,Lachmann:Jablonka:1996}). The ES switching rate decreases as recombination increases until there is an abrupt shift to an ES rate of zero switching. In addition to the ES sex ratio and switching rate, there are other examples where the genetic parameters of the modifier locus affect the ES value at that locus including dispersal and recombination in changing environments \cite{Carja:Liberman:2014}.

@@

This work was supported by National Science Foundation awards \#1846260 and \#1953223 to J.V.C.

>>>>

\usepackage{setspace}
\usepackage{titling}
\usepackage{xparse}

\usepackage[displaymath, mathlines]{lineno}
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
  \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
  \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
  \renewenvironment{#1}%
     {\linenomath\csname old#1\endcsname}%
     {\csname oldend#1\endcsname\endlinenomath}}%
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
  \patchAmsMathEnvironmentForLineno{#1}%
  \patchAmsMathEnvironmentForLineno{#1*}}%
\AtBeginDocument{%
\patchBothAmsMathEnvironmentsForLineno{equation}%
\patchBothAmsMathEnvironmentsForLineno{align}%
\patchBothAmsMathEnvironmentsForLineno{flalign}%
\patchBothAmsMathEnvironmentsForLineno{alignat}%
\patchBothAmsMathEnvironmentsForLineno{gather}%
\patchBothAmsMathEnvironmentsForLineno{multline}%
}

@@

\newcommand{\der}{\mathop{}\!\mathrm{d}}
\newcommand{\Oh}[1]{O\left(#1\right)}
\newcommand{\ExpectationOperator}{\mathrm{E}}
\NewDocumentCommand \E {d[] d()} {%
  \IfNoValueTF {#1} {%
    \IfNoValueTF {#2} {%
      \mathop{\kern0pt\ExpectationOperator}\nolimits%
    }{%
      \mathop{\kern0pt\ExpectationOperator}\nolimits\left(#2\right)%
    }}{%
    \IfNoValueTF {#2} {%
      \mathop{\kern0pt\ExpectationOperator#1}\nolimits%
    }{%
      \mathop{\kern0pt\ExpectationOperator#1}\nolimits\left(#2\right)%
    }%
  }
}
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\Cov}{Cov}
\newcommand{\Fst}{F_{\mathrm{ST}}}
\newcommand{\mean}[1]{\overline{#1}}
\newcommand{\w}{w}
\newcommand{\ess}[1]{#1^*}
\newcommand{\fixp}[1]{\hat{#1}}
\renewcommand{\vec}[1]{\symbf{#1}}
\newcommand{\rec}{\rho}
\newcommand{\mut}{\mu}
\newcommand{\eig}{\lambda}

@@

The role of evolutionary game theory and the ESS method specifically in evolutionary biology has been at the crux of some of the most important conceptual debates in the field including the role of natural selection and adaptation vis-a-vis other forces \cite[e.g.,][]{MaynardSmith:1978,Gould:Lewontin:1979,Lewontin:1979,Orzack:Sober:1994,Gardner:2017,Kern:Hahn:2018,Jensen:Payseur:2019} and the importance of kin and group selection relative to individual selection
\cite[e.g.,][]{Maynard-Smith:1964,Hamilton:1963,Price:1972:cov,Wilson:Wilson:2007,Leigh:2010,Akcay:VanCleve:2012,West:Griffin:2007,Gardner:Grafen:2009,Nowak:Tarnita:2010,Abbot:Abe:2011,Allen:Nowak:2013,Birch:2014,Birch:2017,Nowak:McAvoy:2017}. These debates might be captured in part by the following two questions: \begin{ilnum} \item \label{q:I} How do ESS models that focus on individual fitness capture the effects of kin selection or group selection? \item \label{q:II} How does an ESS account for recombination, mutation, genetic drift and other evolutionary forces? \end{ilnum} In this work, I will describe how conceptual advances in evolutionary theory since Maynard Smith and Price \cite{Maynard-Smith:Price:1973} have shed light on both of these questions and have led to a broader understanding of the ESS method and a more integrative framework for understanding how multiple evolutionary forces in complex populations can lead to diverse phenotypes.

Question \ref{q:I} derives in part from how Maynard Smith introduced his work on ESSs; he argued that an ESS could provide an explanation for the evolution of behaviors where ``selection acts entirely at the individual level, but in which the success of any particular strategy depends on what strategies are adopted by other members of the population'' \cite[p. 210]{Maynard-Smith:1974} and is not ``due to group or species selection'' \cite[p. 15]{Maynard-Smith:Price:1973} or selection on ``close relatives'' \cite[p. 210]{Maynard-Smith:1974} as might be the case for kin selection \cite{Hamilton:1964}. In setting up this dichotomy, Maynard Smith implies that the ESS method may not apply when kin or group selection are involved and that kin or group selection models might lead to different results when applied to the same biological scenario. As I will describe below, evolutionary theorists have since realized (including Maynard Smith himself \cite[p. 33]{MaynardSmith:1978}) that the ESS method is really orthogonal to issues of individual vs. group vs. kin selection and instead captures in what sense evolution via natural selection optimizes fitness given a specific measure of fitness. Issues of individual vs. group vs. kin selection are about the ``units'' or ``levels'' at which selection acts and how fitness should be measured to account for selection at those levels. Lewontin also hints at the levels of selection question in his 1961 game theory paper when he discusses the relative merits of individual-level measures of fitness like intrinsic or Malthusian growth rate $r$ versus population-level measures like mean fitness $\mean{w}$ \cite[pp. 400-401]{Lewontin:1961} as analogs of utility. The levels of selection question became a major topic of study in evolutionary theory \cite[e.g.,][]{Lewontin:1970,Dawkins:1982,Wilson:Sober:1989,Maynard-Smith:Szathmary:1995,Wilson:1997,Michod:1999,Michod:2006,Szathmary:2015} and philosophy of biology \cite{Hull:1980,Brandon:1982,Damuth:Heisler:1988,Lloyd:1992,Lloyd:1994,Sober:Wilson:1994,Okasha:2006,Okasha:2016} and continues to generate substantial research \cite[e.g.,][]{Black:Bourrat:2020,Cooney:Mori:2022,Veit:2022}. For the purpose of explaining how an ESS can capture selection at multiple levels and among relatives or kin, I will argue below that the right measure of utility is the ``lineage fitness'' \cite{Lehmann:Alger:2015,Akcay:VanCleve:2016,Lehmann:Mullon:2016} of a mutant allele at a single genetic locus, which can be shown to be functionally equivalent to a measure of inclusive fitness \cite{Lehmann:Mullon:2016,Lehmann:Rousset:2020}.

@@

The definition of an ESS is relatively simple and only involves a measure of fitness, a set of phenotypes, and a stability condition. Let $\w(x,y)$ be the fitness that an individual with phenotype or strategy $x$ obtains when interacting with an individual with phenotype or strategy $y$. The phenotypes can be drawn from a set of discrete or continous set of values $X$ that represent a single trait like body size or propensity to cooperate. Fitness $\w$ for the moment is simply the survival rate (i.e., viability selection) but as we'll see we can define much more general fitness measures. The stability condition \cite{Maynard-Smith:Price:1973,Maynard-Smith:1974} says that a phenotype $\ess{x}$ is an ESS, or is evolutionarily stable (ES), when no alternative strategy $y$ can improve upon the fitness that $\ess{x}$ receives when interacting with itself:
\begin{subequations}
  \label{eq:ess:def}
\begin{equation}
  \label{eq:ess:def:a}
  \w(\ess{x}, \ess{x}) \ge \w(y, \ess{x})
\end{equation}
for all phenotypes $y$. In the case that the strategy $y$ gets the same fitness interacting with $\ess{x}$ as $\ess{x}$ does with itself (equation \eqref{eq:ess:def:a} holds as an equality) then $\ess{x}$ must do better interacting with $y$ than $y$ does with itself:
\begin{equation}
  \label{eq:ess:def:b}
  \w(\ess{x}, y) > \w(y, y)
\end{equation}
\end{subequations}
The two conditions in \eqref{eq:ess:def} are equivalent to requiring that phenotype $\ess{x}$ receives higher average fitness than any alternative phenotype $y$ when $y$ is rare in the population \cite{Maynard-Smith:1974,Bishop:Cannings:1976}. If condition \eqref{eq:ess:def} holds as a strict inequality when $y \ne \ess{x}$, then $\ess{x}$ is a ``strict'' ESS.

@@

is the frequency of phenotype 1 among mutant individuals once the invasion process has stabilized. Note that since the mutant allele is rare, mutant individuals can have only one mutant chromosome; the other chromosome has no mutant alleles at locus $A$ and remains at the internal equilibrium frequency $\fixp{p}_{kl}$. A sufficient condition for the equilibrium with phenotype $\fixp{x}$ to be externally stable is that $\eig < 1$ or from equation \eqref{eq:lambda:twolocus}

@@

  w(x_{\mut}, \fixp{x}) < w(\fixp{x}, \fixp{x})

@@

for all possible mutant $A_{\mut}$ that generate mutant phenotype $x_{\mut}$. This condition is in fact the condition for a strict ESS from \eqref{eq:ess:def} and can be used to prove that the following result for randomly mating diploid populations with two-locus genetic systems \cite{Eshel:Feldman:1984,Eshel:1996,Eshel:Feldman:1998}:

@@

\section{Invasion fitness and ESSs}

@@

While Result \ref{res:twolocus} emphasizes the role of ESSs and phenotypic evolution in the long-term process, the external stability approach used to prove \ref{res:twolocus} reveals a more general viewpoint where ESSs can depend in important ways on the genetic and demographic parameters of a population. External stability analysis assumes that the population is at some genetic equilibrium where new mutant alleles can invade at low frequency and looks for a genetic equilibrium and its resulting phenotype that resists invasion by any mutant allele within a set of phenotypic effects. Before the mathematical machinery of external stability was developed, Hamilton appreciated the conceptual importance of uninvadable phenotypes in his study of the sex ratio, and he called sex ratios that resisted invasion ``unbeatable'' phenotypes \cite{Hamilton:1967}. As we saw above, determining whether a phenotype can be unbeatable and its genetic equilibria externally stable involves calculating the magnitude of the eigenvalue $\eig$ associated with the linear dynamical system that approximates the dynamics of the mutant allele frequency when rare. The magnitude of the eigenvalue $\eig$ measures the geometric growth rate (e.g., equation \eqref{eq:lambda:twolocus}) of all mutant genotypes in a population of resident genotypes. If there are multiple possible genotypes with the mutant allele, as is the case for the two-locus model in section \ref{sec:long-term} where there may be many possible alleles at the non-mutant locus, then mutant genotype fractions during the invasion process are given by the normalized right eigenvector of the linear system $\vec{v}$. Thus, the $\eig$ really measures the fitness of the whole mutant genetic lineage. The growth rate $\eig$ is often called the ``invasion fitness'' \cite{Metz:Nisbet:1992,Heino:Metz:1998}\footnote{When invasion occurs at more complex attractors like cycles, the ``invasion exponent'' is used, which is $\log |\eig|$ \cite{Rand:Wilson:1994}} and sometimes called the ``lineage fitness'' \cite{Lehmann:Alger:2015,Akcay:VanCleve:2016,Lehmann:Mullon:2016,Lehmann:Rousset:2020}. In effect, the invasion or lineage fitness measures the number of offspring copies that a mutant allele produces among all the genotypes in the mutant lineage \cite{Lehmann:Mullon:2016}. Assuming that $\eig >0$, the phenotype $\ess{x}$ is unbeatable whenever
\begin{equation}
  \label{eq:unbeatable}
  \eig(y, \ess{x}) < 1 \qquad \text{for all } y \ne x \: .
\end{equation}
Regardless of the complexity of the genetics, demography, or ecology of a population, we can in principle perform an external stability analysis\footnote{This is also referred to as an ``evolutionary invasion analysis''.} and use invasion fitness to determine whether there are unbeatable phenotypes.

It should already be evident given Result \ref{res:twolocus} that the unbeatable phenotypes obtained via external stability are closely connected to phenotypes obtained via an ESS analysis using condition \eqref{eq:ess:def}. In the general case for an arbitrary genetic and demographic scenario, we assume only that we have the invasion fitness $\eig(y, \ess{x})$ of a rare mutant $y$ in the population that otherwise has mean phenotype $\ess{x}$. Starting with the condition for a strict ESS, $w(\ess{x}, \ess{x}) > w(y, \ess{x})$ for $y \ne \ess{x}$ from \eqref{eq:ess:def:a}, we convert that condition to one in terms of ``relative fitness'' by dividing by the mean population fitness $w(\ess{x}, \ess{x})$ to obtain
\begin{equation}
  \label{eq:strict:relative}
  \frac{w(y, \ess{x})}{w(\ess{x}, \ess{x})} = \omega(y, \ess{x}) < 1 \qquad \text{for all } y \ne x
\end{equation}
where $\omega(y, \ess{x})$ is the relative fitness of a rare mutant population with mean phenotype $y$ in a population with mean phenotype $\ess{x}$. The strict ESS condition in \eqref{eq:strict:relative} using relative fitness is identical to the condition for uninvadability in condition \eqref{eq:unbeatable} using invasion fitness, which means that unbeatable or uninvadable phentoypes are strict ESSs\footnote{The relationship between invasion fitness and non-strict ESSs is more complex and depends on specific details of the genetics of the population. For example, see the section entitled ``Long-term Stability of an ESS'' in \cite{Eshel:Feldman:1998}}.

Since we can in principle calculate invasion fitness for populations with complex genetics and demography, we can obtain strict ESSs in those populations even though Result \ref{res:twolocus} may not apply. However, those ESSs will almost certainly depend on genetic and demographic parameters of the population as the invasion fitness captures the effect of genetic and demographic processes on the growth of invading mutant lineages\footnote{see also the discussion of ``evolutionary genetic stability'' in \cite[pp. 505--506]{Eshel:1996}}. The fact these ESSs incorporate genetic and demographic forces alongside natural selection shows that the ESS method is a more general tool than potentially originally envisioned by Maynard Smith and other proponents of the phentoypic gambit. Below, we will describe two cases that capture this flexibility: group- and class-structured populations with kin selection and populations in variable environments with bet-hedging phenotypes.

\section{Group structure, inclusive fitness, and ESS}

\section{Bet-hedging, recombination, and ESS}

A more recent example involves the evolution of bet-hedging where organisms alternative between different phenotypes with a specific probability or rate when the environmental cycles between different conditions \cite{Seger:Brockmann:1987,Slatkin:1974}. These phenotypic shifts can be due to genetic mutations \cite{Leigh:1970,Ishii:Matsuda:1989} or switching between different epigenetic states  \cite{Kussell:Leibler:2005,Salathe:VanCleve:2009} where control of the mutation or switching rate is determined by a ``modifier'' locus separate for the genetic or epigenetic locus controlling the phenotype. Due to the the fitness effects of a changing environment, this scenario does not fit neatly into the conditions required for Result \ref{res:twolocus}. In fact, an analysis of the external stability of the switching/mutation rate shows that the switching rate that is ES depends on the recombination rate between phenotypic locus and the modifier locus \cite{Liberman:VanCleve:2011}. For example, Figure \ref{fig:pip:switching}, shows pairwise invasibility plots for switching rate with different recombination rates and given weakly asymmetric selection in two different environments that alternate every $50$ generations. White regions are where mutant alleles with a switching rate on the veritcal axis can invade a population with a resident rate for the rest of the population given by the horizontal axis; black regions are when such a mutant cannot invade. The plots show that a rate of $~1/50$ is ES when the recombination rate is zero (similar to older results in \cite{Leigh:1970,Lachmann:Jablonka:1996}). The ES switching rate decreases as recombination increases until there is an abrupt shift to an ES rate of zero switching. In addition to the ES sex ratio and switching rate, there are other examples where the genetic parameters of the modifier locus affect the ES value at that locus including dispersal and recombination in changing environments \cite{Carja:Liberman:2014}.

@@

This work was supported by National Science Foundation awards \#1846260 and \#1953223.
